### 2023-03-09 paper.tex
rm Ensembl gene ID
inserted missing sum sign
add bookmarks pkg to fix PDF ToC
Added extra toc entries
moved hrule
start re-write

Lukas P.~M. Kremer\textsuperscript{1,2,*}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2}, Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]
\textsuperscript{2} Molecular Neurobiology Dept., German Cancer Research Center (DKFZ), Heidelberg, Germany}
\vspace{1.5ex}

June 2022
\vspace{5.5ex}

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution. Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Typically, this is simply done by dividing the genome into large tiles and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging. We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number. We also show that our method of finding variably methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality required to work with scBS data.
\end{minipage}
\vspace{6.5ex}

}]

@@

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample. Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up. 
\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}

@@

In the present paper, we discuss strategies to analyze scBS data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

\begin{figure*}[t]
    \hspace{2.5cm}A\\[-7ex]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
    \end{center}
    \hspace{2.5cm}B\\[-3ex]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
    \end{center}
    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells. \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained. To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
    \label{fig:smoothres}
\end{figure*}

@@

This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile. To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA. After PCA, one can proceed using approaches known from scRNA-seq.

@@

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval. In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown. The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

@@

How should one obtain the ensemble average (the curved line in Fig.\ \ref{fig:smoothres}B)? A simple approach to get a value for a specific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated. However, especially when only few cells offer coverage, these averages will be very noisy. Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood. The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; for the examples presented here we used 1000~bp.

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
    \caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{fig:smoothres}B. Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold. Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
    \label{vmr}
\end{figure}

Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells. For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}. In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
Only the latter regions are of value for our goal of quantitating dissimilarity between cells. We call these the variably methylated regions (VMRs).

In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

Therefore, we propose the following approach (Fig. \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B. Then, calculate for each window the variance of these values over all cells. Select, say, the top 2\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are separated by only a small gap, merge them into one larger VMR. Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.

@@

@@

\section{Benchmarks}

@@

\begin{figure}
    \begin{center}
    A\hspace{.9\columnwidth}~\\
    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
    B\hspace{.9\columnwidth}~\\
    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
    C\hspace{.9\columnwidth}~\\
    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis). Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(B)} Same as in (A), but using the shrunken mean of residuals as a measure of methylation.
    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in orange in A \& B). Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression). The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
    Data from \citet{argelaguet2019gastru}.}
    \label{figure:correlation}
\end{figure}

@@

\begin{figure*}
    \begin{center}
        A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
        \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
        \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
        C\hspace*{.7\textwidth}~\\
        \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
        D\hspace*{.7\textwidth}~\\
        \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
        \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb tiles) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or shrunken means of residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fractions of 100~kb tiles (left) or shrunken means of residuals of VMRs (right). Data from \citet{luo2017single}.}
        \label{figure:score}
    \end{center}
\end{figure*}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken means of residuals as a measure of methylation. 
For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.

@@

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
% why not just plot PC1 vs. PC2 here? % because people will want to see the UMAP. -S
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
        These cells can be discarded with \texttt{scbs filter}.
        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
    }
    \label{workflow}
\end{figure*}

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command. This produces a BED file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$). For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of residuals were obtained with \texttt{scbs matrix}. Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded. To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in. 
For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section. To visualize data, UMAP was performed on the final 15 principal components.

>>>>

Lukas P.~M. Kremer\textsuperscript{1,2,*}, Martina Braun\textsuperscript{2}, Leonie Küchenhoff\textsuperscript{1,2}, Santiago Cerrizuela\textsuperscript{2},\\Ana Martin-Villalba\textsuperscript{2}, Simon Anders\textsuperscript{1,*}\\[1ex]
{\footnotesize \textsuperscript{1} BioQuant Centre, University of Heidelberg, Heidelberg, Germany\\[-.5ex]
\textsuperscript{2} Molecular Neurobiology Dept., German Cancer Research Center (DKFZ), Heidelberg, Germany}
\vspace{1.5ex}

January 2023
\vspace{5.5ex}

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution.
Analyzing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability.
Typically, this is simply done by dividing the genome into large tiles and averaging in these.
We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging.
We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number.
We also implement an approach to detect differentially methylated regions (DMRs) between groups of cells.
Finally, we present a software tool, \texttt{scbs}, that implements these methods and offers further functionality required to work with scBS data.
\end{minipage}
\vspace{6.5ex}
}]

@@

Sequencing-based assays with single-cell resolution have offered new means to understand the differences between the cells making up a sample.
Single-cell RNA sequencing (scRNA-seq) techniques have matured at great pace in recent years, with well developed analysis methodology, and methods to study epigenetics at single-cell resolution are rapidly catching up.

\blfootnote{\hspace{-.5cm}\raggedright\textsuperscript{*} e-mail:  \texttt{l.kremer@dkfz-heidelberg.de} and\\ \hspace{1cm}\texttt{simon.anders@bioquant.uni-heidelberg.de}}

@@

In the present paper, we discuss strategies to analyze single-cell bisulfite-sequencing (scBS) data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
    \end{center}
    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:}
    \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
Two cells cover differing parts of the interval with one read each.
If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
    \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
    \label{fig:smoothres}
\end{figure*}

@@

This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for PCA.
A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g. 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell.
Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1.
This matrix is now subjected to PCA.
After PCA, one can proceed with dimensionality reduction and clustering approaches known from scRNA-seq.

@@

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval.
In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown.
The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having higher methylation in the interval.
However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

@@

How should one obtain the ensemble average (the curved line in Fig.~\ref{fig:smoothres}B)? A simple approach to get a value for a specific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated.
However, especially when only few cells offer coverage, these averages will be very noisy.
Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood.
The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; for the examples presented here we used 1000~bp.

% remove or shorten this section cause it's not really novel, methods like this are already implemented in e.g. the "missMDA" R package
\subsection{PCA with imputation}

@@

\subsection{Finding variably methylated regions (VMRs)}

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
    \caption{\small \textbf{Finding variably methylated regions (VMRs):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
    \label{vmr}
\end{figure}

Typically, some regions in a chromosome will have very similar methylation status in all cells, while other regions show variability in methylation across cells.
For instance, it is long known that CpG-rich promoters of housekeeping genes are unmethylated, and that a large proportion of the remaining genome is highly methylated regardless of cell type \citep{bird1986cpg}.
In contrast, DNA methylation at certain genomic features such as enhancers is more dynamic \citep{argelaguet2019gastru}, and thus more variable across cells.
Only the latter regions are of value for our goal of quantitating dissimilarity between cells.
We call these the variably methylated regions (VMRs). % maybe mention the parallel to highly variable genes in scRNAseq?

In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile.
Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpG sites that are equal in all cells, when averaging over all the CpG sites in the tile.

Therefore, we propose the following approach (Fig.~\ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size.
Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Fig.~\ref{fig:smoothres}B.
Then, calculate for each window the variance of these values over all cells.
Select, say, the top 2\% windows with the highest variances and mark them as VMRs.
Wherever thus marked windows overlap or are separated by only a small gap, merge them into one larger VMR.
Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.

@@

\subsection{Finding differentially methylated regions (DMRs)}

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
    \caption{\small \textbf{Identification of differentially methylated regions (DMRs):}
    The chromosome is divided into windows overlapping by a small and fixed size.
    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~ \ref{fig:smoothres}B.
    Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
    Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
    Windows exceeding either threshold are merged if they overlap, yielding the ``differentially methylated regions``.}
    \label{dmr}
\end{figure}

A common task in the analysis of bulk bisulfite-sequencing data is the detection of differentially methylated regions (DMRs) between conditions, tissues, or cell types \citep{Hebestreit2013, dmrseq}.
As DNA methylation affects gene expression, DMRs can provide insights into the unique epigenetic and gene regulatory characteristics of cell types.
However, to date no approach to detect DMRs in scBS data has been reported.
To enable DMR detection in scBS data, we thus developed an approach that detects DMRs of variable size between two groups of cells, and controls the false discovery rate (Fig.~3):

Similar to the previously described VMR identification approach (Fig.~\ref{vmr}), divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window as described (Fig.~\ref{residuals}b).
Next, instead of calculating the variance, obtain the t-statistic % Welch t-test
as a measure of differential methylation between the two cell groups.
Mark the windows with the most extreme t-statistics (e.g. 2\% upper and lower tails) as DMRs.
Distinguishing between high and low t-statistics, merge any overlapping marked windows into larger DMRs.
Re-calculate the t-statistic for each larger DMR.

To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data.
The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.

\subsection{???}  % guess we have to rewrite this closing paragraph a little bit, to include both DMRs and VMRs

@@

\section{Benchmarks} % re-name this to something more general? Application and benchmarks or something?

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (mean shrunken residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{figure:correlation}
\end{figure}

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\columnwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
    \textbf{(A)} Mean neighbor score obtained when analyzing data from \citet{kremer_scnmt} with different methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation fractions, based on \citep{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(B)}.
    \textbf{(C, D)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (C) or conventional methods based on genome tiling (D).}
    \label{figure:score}
\end{figure}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}.
This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken means of residuals as a measure of methylation.

For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.

@@

Of note, our results show that even few VMRs (here, the 1000 VMRs that have read coverage in the highest number of cells) already contain sufficient information to distinguish neuronal subtypes.
In contrast, the neighbor count score for the top 1000 genomic 100~kb tiles is much lower.
To visualize the different degrees of separation, we further reduced the PC space from both analyses with UMAP, showing that some neuronal subtypes such as mL5-1 and mDL-3 are separated from the other subtypes when using 1000 VMRs (Figure \ref{figure:score}C).
Again, the difference is even more pronounced in the sub-sampled analysis with only 500 cells (Figure \ref{figure:score}D).

\subsection{Identification of biologically relevant DMRs}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{part_martina/NbvsOlig_DMRs.png}
    \end{center}
    \caption{\small \textbf{identified neuroblast and oligodendrocyte DMRs have a putative regulatory role:} \textbf{(A)} Difference in methylation between oligodendrocytes and neuroblasts of DMRs plotted against the number of cells contained in a DMR.
Colored by p-value.
\textbf{(B)} Mean methylation of cells in the DMR covering MBP.
Line indicates median.
\textbf{(C)} Mean methylation of reads at individual sites along the DMR covering MBP.
\textbf{(D)} GREAT \citet{mclean2010great} was used for GO-term analysis.
The left side shows GO-terms for genes near the 3000 most significant neuroblast DMRs sorted by GO-enrichment p-value and t-statistic (green arrow) and the right side for oligodendrocyte DMRs (yellow arrow) respectively.
    }
    \label{dmr_oligVSnb}
\end{figure*}

As a demonstration and to assess the ability of the presented approach to detect relevant DMRs, the method was used to find regions differing in CpG methylation between vSVZ neuroblasts and oligodendrocytes.
Figure 7A gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
A low p-value is maintained, although the methylation difference decreases if the number of cells in the DMR increases.

An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Figure 5B-C).
Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Figure 7B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Figure 5C).
Low methylation at this locus supports the expression of Mbp.

Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Figure 5D).
On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
Therefore, DMRs occur in the proximity of genes associated with oligodendrocyte or neuroblast function, respectively and are implied to regulate gene expression and contribute to fundamental cell type-specific function, which is supported by the repressive effect of CG methylation on gene expression.

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
        These cells can be discarded with \texttt{scbs filter}.
        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
        Identified groups of cells (e.~g.
clusters) can then be scanned for DMRs with \texttt{scbs diff}.
    }
    \label{workflow}
\end{figure*}

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
Similarly, the user can identify DMRs after quality control with \texttt{scbs diff} using a text file listing the cells’ group assignment as input.
The command produces a BED file listing the coordinates of the determined DMRs together with the t- statistic, the corresponding cell group, and the adjusted p-value.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of residuals were obtained with \texttt{scbs matrix}.
Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.

For dimensionality reduction, missing values were imputed with the iterative PCA described in the previous section.
To visualize data, UMAP was performed on the final 15 principal components.
